---
title: "Fast disconnected ICAR in Stan"
description: |
  Using the segment feature in Stan to implement the ICAR model
author:
  - name: Adam Howes
    affiliation: Imperial College London
date: 2020-11-24
source_url: true
preview: fast-disconnected-icar-in-stan.jpg
output:
  distill::distill_article:
    toc: true
    toc_depth: 2
    self_contained: false
bibliography: citations.bib
draft: TRUE
---

```{r}
knitr::opts_chunk$set(eval = FALSE)
```

*Note: actually, Mitzi already got to this – see [here](https://github.com/stan-dev/example-models/blob/master/knitr/car-iar-poisson/bym2_islands.stan) for her version!*

## Introduction

@morris2019bayesian provide a (more) efficient implementation of the intrinsic conditional auto-regressive (ICAR) model in Stan.
In this post I extend their approach to the case when the adjacency graph is disconnected, using [Section 8.2. of the Stan User’s Guide](https://mc-stan.org/docs/2_25/stan-users-guide/ragged-data-structs-section.html) for indexing.
Their paper follows from the [Spatial Models in Stan: Intrinsic Auto-Regressive Models for Areal Data](https://mc-stan.org/users/documentation/case-studies/icar_stan.html) case study which I’d recommend having a look at.

## Background

Consider areas $\{A_i\}_{i = 1, \ldots, n}$ with corresponding spatial random effects $\boldsymbol{\phi} = (\phi_1, \ldots, \phi_n)^\top$ distributed according to the zero-mean intrinsic Gaussian Markov random field
$$
\boldsymbol{\phi} \sim \mathcal{N}(\mathbf{0}, \tau_\phi^{-1} \mathbf{R}^{-}), \label{eq:gmrf},
$$
where $\tau_\phi$ is the precision and $\mathbf{R}^{-}$ is the pseudo-inverse of the rank-deficient structure matrix $\mathbf{R}$ (such that the precision matrix is given by $\tau_\phi \mathbf{R}$).
One way to specify the structure matrix is by the Laplacian of the adjacency graph $\mathcal{G} = (\mathcal{V}, \mathcal{E})$, with $n$ vertices $v \in \mathcal{V}$ corresponding to each area and edges $e \in \mathcal{E}$ between vertices $i$ and $j$ when areas are adjacent, denoted $i \sim j$.
This results in an improper density
$$
p(\boldsymbol{\phi}) \propto \exp \left( -\frac{\tau_\phi}{2} \boldsymbol{\phi}^\top \mathbf{R} \boldsymbol{\phi} \right),
$$
so called because it is invariant to certain linear transformations of $\boldsymbol{\phi}$.

Two important issues when working with these types of models are:

* *Scaling the structure matrix* In order for $\tau_\phi$ to have a consistent interpretation across models, the structure matrix $\mathbf{R}$ (which is a function of the adjacency graph) requires scaling. This "consistent interpretation across models" is useful when you are setting priors, among other things. When $\tau_\phi = 1$ we would like the "typical marginal variance" to be one as well. This requires scaling the structure matrix, say replacing it with a scaled version $\mathbf{R}^\star$.

* *Constraining the spatial random effects* So that the random effect can be identified we place constraints on $\boldsymbol{\phi}$. It is standard (for example in `R-INLA`) to use sum-to-zero constraints of the form $\sum_i \phi_i = 0$.

Particular care is required when $\mathcal{G}$ is disconnected (a graph is disconnected when there exist areas which can’t be reached following the vertices of the adjacency graph.)
For this reason @morris2019bayesian consider only connected graphs, stating that the indexing required in the disconnected case would complicate the code for presentation purposes.
In some settings the graph is disconnected and there is little that can be done about it.
Hopefully this post provides a starting point.

Recommendations for dealing scaling and constraints in the disconnected case are given by @freni2018note.
To summarise, they firstly recommend to scale local smoothing in each connected component independently so that τϕ has a consistent interpretation within the model.
Secondly, they recommend to place a sum-to-zero constraint on each connected component made up of at least two areas.
This results in “singletons” having independent zero-mean Gaussian random effects (rather than constraining them to be identically zero which is another possible choice).

## Example implementation

```{r}
library(bsae)
library(dplyr)
library(ggplot2)
library(rstan)
```

Malawi (if you attend Imperial’s HIV inference group meetings you’ll know that this is the only country which exists) has 28 districts and one of them, [Likoma](https://en.wikipedia.org/wiki/Likoma_Island) is an island -- or singleton.

`bsae` is a personal package I am working on which contains the [simple features](https://r-spatial.github.io/sf/index.html) dataframe `mw` which has data about HIV cases `mw$y` and sample sizes `mw$n_obs` in Malawi from a particular survey.
A neighbours object can be extracted from the geometry `mw$geometry` giving:

```{r}
nb <- sf_to_nb(mw)
nb
```

As you can see there is one region, Likoma, with no links.
`nb` can be used to create the unscaled structure matrix:

```{r}
R <- nb_to_precision(nb)
plot_matrix(R)
```

In the second row and column, all of the values are zero. In particular, the entry `R[2, 2]` is zero giving $\phi_2$ infinite variance according to the unscaled prior!
Using `scale_grmf_precision` replaces this value by one such that $\phi_2 \sim \mathcal{N}(0, \tau_\phi^{-1})$ as well as scaling the submatrix corresponding to the mainland such that the mean of the marginal variances is one.

```{r}
R_scaled <- scale_gmrf_precision(R)
plot_matrix(R_scaled$Q)
```

The matrix `R_scaled$Q` is very similar to `R`, just with a different value at `[2, 2]` and the rest of the matrix rescaled.

I will now fit a Bayesian hierarchical model to the case data, using a ICAR spatial random effect model following @morris2019bayesian, making some adjustments to deal with the disconnected graph.

To pass Stan information about the connected components I use three entries in the data block:

* `nc`, the number of connected components, `nc = 2` in this case
* `group_sizes`, a vector of length `nc` giving the number of areas within each connected component, `group_sizes = c(27, 1)` here
* `scales`, a vector of length `n` used to rescale the GMRF

The Stan code for the model is:

```{stan output.var = "fast_disconnected"}
data {
  int<lower=1> n; // Number of regions
  int y[n];       // Vector of responses
  int m[n];       // Vector of sample sizes
  
  // Data structure for connected components input
  int<lower=1> nc;
  int group_sizes[nc];    
  vector[n] scales;
  
  // Data structure for graph input
  int<lower=1> n_edges;
  int<lower=1, upper=n> node1[n_edges];
  int<lower=1, upper=n> node2[n_edges];
}

parameters {
  real beta_0;             // Intercept
  vector[n] u;             // Unscaled spatial effects
  real<lower=0> sigma_phi; // Standard deviation of spatial effects
}

transformed parameters {
  real tau_phi = 1 / sigma_phi^2; // Precision of spatial effects
}

model {
  // ICAR model
  int pos;
  pos = 1;
  for (k in 1:nc) {
  
    if(group_sizes[k] == 1){
      // Independent unit normal prior on singletons
      segment(u, pos, group_sizes[k]) ~ normal(0, 1);
    }
    
    if(group_sizes[k] > 1){
      // Soft sum-to-zero constraint on 
      // each non-singleton connected component
      sum(segment(u, pos, group_sizes[k])) ~ normal(0, 0.001 * group_sizes[k]);
    }
    
    pos = pos + group_sizes[k];
  }
  // Note that the below increment doesn't feature singletons
  target += -0.5 * dot_self(u[node1] - u[node2]);
  
  // The rest of the model
  y ~ binomial_logit(m, beta_0 + sigma_phi * sqrt(1 ./ scales) .* u);
  beta_0 ~ normal(-2, 1);
  sigma_phi ~ normal(0, 2.5);
}

generated quantities {
  vector[n] phi = sigma_phi * sqrt(1 ./ scales) .* u;
  vector[n] rho = inv_logit(beta_0 + phi);
}
```

The most interesting part is the ICAR prior in the model block.

I have used the segment operation to implement a ragged data structure (the number of areas in each connected component) following the [Stan User’s Guide](https://mc-stan.org/docs/2_25/stan-users-guide/ragged-data-structs-section.html).
Because `u` is arranged with the connected components grouped together, `segment(u, 1, group_sizes[1])` gives all of the mainland and `segment(u, 27, group_sizes[2])` gives Likoma.
The expression

```{stan output.var = "tmp", eval=FALSE}
sum(segment(u, pos, group_sizes[k])) ~ normal(0, 0.001 * group_sizes[k])
```

specifies that the mean of the segment is approximately zero.
@morris2019bayesian find that this type of “soft constraint” works better than a hard constraint where the sum must be identically zero. 

One complication is that using this Stan code requires reordering the data so that the connected components are grouped together.
In the case of `mw` this just involves moving Likoma from the second row to the last, though in general there may be more shuffling.

```{r}
comp <- spdep::n.comp.nb(nb)
comp
```

```{r}
ordered_mw <- mw %>%
  mutate(comp.id = comp$comp.id) %>%
  arrange(comp.id)
```

I create the graph inputs as follows:

```{r}
ordered_nb <- sf_to_nb(ordered_mw)
ordered_g <- nb_to_graph(ordered_nb)
```

And the three inputs used to specify the connected components are:

```{r}
ordered_comp <- spdep::n.comp.nb(ordered_nb)
ordered_comp$nc

rle <- rle(ordered_comp$comp.id) #' Run length encoding 
rle

ordered_scales <- ordered_nb %>%
  nb_to_precision() %>%
  scale_gmrf_precision()

vec_scales <- ordered_scales$scales[ordered_comp$comp.id]
vec_scales
```

`vec_scales` contains the scaling constant 0.7171335 for the mainland and the value 1 for Likoma.
In some sense, really the scaling constant for the island should be Inf, but as you can see in the Stan code we are treating it with a special case (and so don’t want to do alter the scaling).

Now I can pass the data to Stan and start sampling:

```{r results='hide'}
#' A quirk of the data I'm using is that counts might not be integers
#' This is because of survey weighting

dat <- list(
  n = nrow(ordered_mw),
  y = round(ordered_mw$y),  
  m = round(ordered_mw$n_obs),
  nc = comp$nc,
  group_sizes = rle$lengths,
  scales = vec_scales,
  n_edges = ordered_g$n_edges,
  node1 = ordered_g$node1,
  node2 = ordered_g$node2
)

stan_fit <- rstan::sampling(fast_disconnected, data = dat, warmup = 500, iter = 1000)
out <- rstan::extract(stan_fit)
```

```{r}
rstan::summary(stan_fit)$summary[c("beta_0", "phi[28]"), ]
```

There is more exploring to do here!

## Conclusion

To test if this worked I should evaluate it in comparison to a version fitted with `R-INLA`.
Along the lines of:

```{r eval=FALSE}
dat <- list(id = 1:nrow(ordered_mw),
             y = round(ordered_mw$y),
             m = round(ordered_mw$n_obs))

tau_prior <- list(prec = list(prior = "logtnormal", param = c(0, 1 / 2.5^2),
                              initial = 0, fixed = FALSE))

formula <- y ~ 1 + f(id,
                     model = "besag",
                     graph = nb,
                     scale.model = TRUE,
                     constr = TRUE,
                     hyper = tau_prior)

inla_fit <- INLA::inla(formula,
                       family = "binomial",
                       control.family = list(control.link = list(model = "logit")),
                       data = dat,
                       Ntrials = m,
                       control.predictor = list(compute = TRUE, link = 1),
                       control.compute = list(dic = TRUE, waic = TRUE,
                                              cpo = TRUE, config = TRUE))
```

That being said, I have found it often tricky to get Stan and `R-INLA` to agree!
